---
title: "Comprehensive Analysis of DNA Methylation Using the Illumina HumanMethylation450 BeadChip Platform"
subtitle: "DNA & RNA Dynamics final project, Team 1. Repository: [Parganini/DRD-Final-Project-Team1](https://github.com/Parganini/DRD-Final-Project-Team1)"
author: "Fatemeh Bayat, Tamime Eter, Seyed Parsa Golshani, Amirhossein Heidari, Mahmoud Ahmed Mohamed Mohamed Ibrahim, Hector Eduardo Parga Najera, Bita Rahmani, Shayan Zanjaniha"
date: "2026-06"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,
                      fig.align="center", dpi=300, fig.retina=2)
knitr::opts_knit$set(root.dir=normalizePath("."))
```

# Introduction


Team 1 assignments:

- Address for the fluorescence check: `42796479`
- Detection p-value threshold: `0.05`
- Normalization method: `preprocessNoob`
- Differential methylation test: `t-test`

# Step 00. Basic Preparation

Clean the R environment

```{r clean-environment}
rm(list=ls())
```

Load the packages required for this analysis

```{r load-packages}
library(minfi)
library(qqman)
library(gplots)
library(IlluminaHumanMethylation450kmanifest)
```

Define the folders used by the project

```{r define-folders}
baseDir <- "data/raw"
rdsDir <- "results/rds"
tablesDir <- "results/tables"
figuresDir <- "results/figures"
```

# Step 01. Data Import and RGChannelSet Creation

```{r step01-load-data}
list.files("./data/raw/")
SampleSheet <- read.csv("./data/raw/SampleSheet_Report_II.csv", header = TRUE)
SampleSheet

targets <- read.metharray.sheet("./data/raw")
targets

RGset <- read.metharray.exp(targets = targets)
save(RGset, file = file.path(rdsDir, "RGset_Report.RData"))

RGset
str(RGset)
```

The `SampleSheet` contains the metadata for our 8 samples, including group (CTRL/DIS), sex, and Sentrix ID/Position. `read.metharray.sheet()` reads this sheet and automatically attaches the file paths needed to locate the corresponding idat files. `read.metharray.exp()` then loads the raw Red and Green fluorescence intensities for all samples into an `RGChannelSet` object called `RGset`.

# Step 02. Red and Green Fluorescence Channels

```{r step02-extract-channels}
Red <- data.frame(getRed(RGset))
dim(Red)
head(Red)

Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)
```

`getRed()` and `getGreen()` extract the raw fluorescence intensity matrices from `RGset`, one row per probe address and one column per sample.

# Step 03. Team 1 Address Check

```{r step03-address-lookup}
TARGET_ADDRESS <- "42796479"

getManifest(RGset)
ProbeInfo_I  <- data.frame(getProbeInfo(RGset, type = "I"))
ProbeInfo_II <- data.frame(getProbeInfo(RGset, type = "II"))

ProbeInfo_I[ProbeInfo_I$AddressA == TARGET_ADDRESS, ]
ProbeInfo_I[ProbeInfo_I$AddressB == TARGET_ADDRESS, ]
ProbeInfo_II[ProbeInfo_II$AddressA == TARGET_ADDRESS, ]

Red_fluorescences   <- Red[rownames(Red) == TARGET_ADDRESS, ]
Green_fluorescences <- Green[rownames(Green) == TARGET_ADDRESS, ]

df_summary_probes <- data.frame(
  Sample             = colnames(Red_fluorescences),
  Red_Fluorescence   = as.numeric(Red_fluorescences[1, ]),
  Green_Fluorescence = as.numeric(Green_fluorescences[1, ]),
  Type               = "I",
  Color              = "Red"
)

print(df_summary_probes)
```

Address `42796479` corresponds to the **AddressA (unmethylated)** position of probe **cg10602367**, a **Type I, Red-channel** probe. Type I probes use a single color channel, so the Green fluorescence values for this address are near background levels across all samples, while Red fluorescence varies between samples — reflecting biological differences in signal intensity at this CpG site.

# Step 04. MethylSet Creation

```{r step04-mset-raw}
MSet.raw <- preprocessRaw(RGset)
MSet.raw
save(MSet.raw, file = file.path(rdsDir, "MSet_raw.RData"))

Meth <- as.matrix(getMeth(MSet.raw))
str(Meth)
head(Meth)

Unmeth <- as.matrix(getUnmeth(MSet.raw))
str(Unmeth)
head(Unmeth)

Meth[rownames(Meth) == "cg10602367",]
Unmeth[rownames(Unmeth) == "cg10602367",]
```

`preprocessRaw()` converts the raw Red/Green intensities in `RGset` into a `MethylSet`, splitting the signal into **Methylated** and **Unmethylated** intensities for each probe. For our probe `cg10602367`, the Unmethylated signal is substantially higher than the Methylated signal across all 8 samples, consistent with the Red-channel Type I design identified in Step 3.

# Step 05. Quality Control

```{r step05-qc-plot}
qc <- getQC(MSet.raw)
qc
plotQC(qc)
```

**Interpretation:** The QC plot shows the median methylated (mMed) and unmethylated (uMed) signal intensities per sample. Samples with higher values in both signals indicate higher quality. Most samples show median values in the 8.6–11.2 range (log2 scale); a few samples (e.g. GSM5319603, GSM5319609) show comparatively lower values, suggesting somewhat lower signal quality.

```{r step05-negative-controls}
NegativeControl <- data.frame(getProbeInfo(RGset, type = "Control"))
table(NegativeControl$Type)
head(NegativeControl)
controlStripPlot(RGset, controls = "NEGATIVE")
```

**Interpretation:** The expected value for these negative controls is around or below log2(10) (which represents the baseline background fluorescence noise of the scanner). Since these control probes are designed with sequences that should not hybridize to human DNA, they solely measure the baseline random noise of the assay. Because all our samples fall within this low-intensity range across both the Red and Green channels, we can confidently confirm that there is no excessive background fluorescence or nonspecific cross-hybridization in our arrays.

```{r step05-detection-pvalues}
Detection_pvalue <- detectionP(RGset)
save(Detection_pvalue, file = file.path(rdsDir, "Detection_pvalue.RData"))

Over_Threshold <- Detection_pvalue > 0.05
table(Over_Threshold)
sum(Over_Threshold)

Total_Failed_Positions <- sum(Over_Threshold)
Total_Detection_Tests <- length(Over_Threshold)
Total_Failed_Percentage <- Total_Failed_Positions / Total_Detection_Tests * 100

failed_probe_message <- sprintf(
  "There are %s failed detection tests with a detection p-value higher than 0.05",
  Total_Failed_Positions
)

Over_Threshold_Per_Sample <- colSums(Over_Threshold)
Over_Threshold_Per_Sample

Over_Threshold_Fraction_Per_Sample <- colMeans(Over_Threshold)
Over_Threshold_Fraction_Per_Sample

failed <- data.frame(
  Sample = names(Over_Threshold_Per_Sample),
  n_Failed_Positions = Over_Threshold_Per_Sample,
  Fraction_Failed_Positions = Over_Threshold_Fraction_Per_Sample,
  Percentage_Failed_Positions = Over_Threshold_Fraction_Per_Sample * 100
)
print(failed)
print(failed_probe_message)
```

**Interpretation:** Using Team 1's detection p-value threshold of 0.05, there are `r Total_Failed_Positions` failed detection tests across all samples, corresponding to `r round(Total_Failed_Percentage, 2)`% of all probe-sample detection tests. Per sample, the number of failed positions ranges from `r min(Over_Threshold_Per_Sample)` to `r max(Over_Threshold_Per_Sample)`, corresponding to `r round(min(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% to `r round(max(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% of positions. Since all samples show a low percentage of failed positions, the overall detection quality is good and all samples can be kept for the next steps.

# Step 06. Raw Beta and M Values

We retrieve beta and M value matrices from the raw MethylSet.

Extract raw beta values and inspect their dimensions and distribution.
```{r step06-beta-raw}
beta <- getBeta(MSet.raw)
class(beta)
dim(beta)
head(beta)
summary(beta)
```

Extract raw M values and inspect their dimensions and distribution.
```{r step06-m-raw}
M <- getM(MSet.raw)
dim(M)
head(M)
summary(M)
```

Save raw beta and M matrices.
```{r step06-save-beta-m}
saveRDS(beta, file=file.path(rdsDir, "beta_raw.rds"))
saveRDS(M, file=file.path(rdsDir, "M_raw.rds"))
```

Recover phenotype data to split samples by group.
```{r step06-phenotype-data}
pheno <- data.frame(pData(RGset))
str(pheno)
table(pheno$Group)
```

Split beta values into CTRL and DIS groups.
```{r step06-split-beta}
beta_CTRL <- beta[, pheno$Group=="CTRL"]
beta_DIS <- beta[, pheno$Group == "DIS"]

dim(beta_CTRL)
dim(beta_DIS)
```

Split M values into CTRL and DIS groups.
```{r step06-split-m}
M_CTRL <- M[, pheno$Group=="CTRL"]
M_DIS <- M[, pheno$Group=="DIS"]

dim(M_CTRL)
dim(M_DIS)
```

Calculate mean beta and M values per CpG within each group.
```{r step06-summary-beta-m}
mean_of_beta_CTRL <- apply(beta_CTRL, 1, mean, na.rm=TRUE)
mean_of_beta_DIS <- apply(beta_DIS, 1, mean, na.rm=TRUE)

mean_of_M_CTRL <- apply(M_CTRL, 1, mean, na.rm=TRUE)
mean_of_M_DIS <- apply(M_DIS, 1, mean, na.rm = TRUE)
```
Remove infinite M values before density estimation.
```{r step06-clean-infinite-m}
mean_of_M_CTRL <- mean_of_M_CTRL[is.finite(mean_of_M_CTRL)]
mean_of_M_DIS <- mean_of_M_DIS[is.finite(mean_of_M_DIS)]
```

Estimate density distributions for raw beta and M means.
```{r step06-density-estimation}
d_mean_of_beta_CTRL <- density(mean_of_beta_CTRL, na.rm=TRUE)
d_mean_of_beta_DIS <- density(mean_of_beta_DIS, na.rm=TRUE)

d_mean_of_M_CTRL <- density(mean_of_M_CTRL, na.rm=TRUE)
d_mean_of_M_DIS <- density(mean_of_M_DIS, na.rm=TRUE)
```

Plot raw beta density distributions by group.
```{r step06-plot-beta-density, message=FALSE, warning=FALSE}
plot(d_mean_of_beta_CTRL, col="blue",
     main="Raw beta values", xlab="Mean beta")

lines(d_mean_of_beta_DIS, col="red")

legend("topright", legend=c("CTRL", "DIS"),
       col=c("blue", "red"), lty=1, bty="n")
```

Plot raw M value density distributions by group.
```{r step06-plot-M-density, message=FALSE, warning=FALSE}
plot(d_mean_of_M_CTRL, col="blue",
     main="Raw M values", xlab="Mean M")

lines(d_mean_of_M_DIS, col="red")

legend("topright", legend=c("CTRL", "DIS"),
       col=c("blue", "red"), lty=1, bty="n")
```

**Interpretation:** The raw beta and M-value density curves for CTRL and DIS samples show broadly similar distributions. Small shifts between the two groups can be observed, but the overall methylation profiles are largely comparable at the raw-data level.

Save summary vectors used for the density plots.
```{r step06-save-summary-vectors}
saveRDS(mean_of_beta_CTRL,
        file=file.path(rdsDir, "mean_of_beta_raw_CTRL.rds"))

saveRDS(mean_of_beta_DIS,
        file=file.path(rdsDir, "mean_of_beta_raw_DIS.rds"))

saveRDS(mean_of_M_CTRL,
        file=file.path(rdsDir, "mean_of_M_raw_CTRL.rds"))

saveRDS(mean_of_M_DIS,
        file=file.path(rdsDir, "mean_of_M_raw_DIS.rds"))
```

# Step 07. Normalization with preprocessNoob

The assigned Team 1 normalization method was `preprocessNoob`. Raw and normalized beta values were compared by Type I and Type II probe chemistry. This comparison is important because Infinium Type I and Type II assays can show systematic differences: Type II beta values tend to be shifted toward intermediate methylation levels and can show different variability from Type I probes. Normalization is used to reduce technical variation before biological interpretation.

Open the raw RGChannelSet and raw beta matrix saved in previous steps.
```{r step07-load-raw-data}
load(file.path(rdsDir, "RGset_Report.RData"))
beta <- readRDS(file.path(rdsDir, "beta_raw.rds"))
```

Inspect the raw RGChannelSet and beta matrix.
```{r step07-inspect-raw-data}
RGset
head(beta)
dim(beta)
```

Retrieve Type I and Type II probe annotations from the manifest.
```{r step07-probe-annotations}
dfI <- data.frame(getProbeInfo(RGset, type="I"))
dfII <- data.frame(getProbeInfo(RGset, type="II"))

dim(dfI)
dim(dfII)
str(dfI)
str(dfII)
head(dfI$Name)
head(dfII$Name)
```

Split raw beta values by probe chemistry.
```{r step07-split-beta-by-chemistry}
beta_I <- beta[rownames(beta) %in% dfI$Name,]
beta_II <- beta[rownames(beta) %in% dfII$Name,]

dim(beta_I)
dim(beta_II)
```

Calculate mean raw beta by probe chemistry.
```{r step07-mean-beta-by-chemistry}
mean_of_beta_I <- apply(beta_I, 1, mean, na.rm=T)
mean_of_beta_II <- apply(beta_II, 1, mean, na.rm=T)
```

Estimate density of raw beta means by probe chemistry.
```{r step07-density-beta-by-chemistry}
d_mean_of_beta_I <- density(mean_of_beta_I, na.rm=T)
d_mean_of_beta_II <- density(mean_of_beta_II, na.rm=T)
```

Calculate and estimate density of raw beta standard deviations.
```{r step07-beta-sd-by-chemistry}
sd_of_beta_I <- apply(beta_I, 1, sd, na.rm=T)
sd_of_beta_II <- apply(beta_II, 1, sd, na.rm=T)

d_sd_of_beta_I <- density(sd_of_beta_I, na.rm=T)
d_sd_of_beta_II <- density(sd_of_beta_II, na.rm=T)
```
Normalize the raw RGChannelSet with preprocessNoob.
```{r step07-normalization}
preprocessNoob_results <- preprocessNoob(RGset)
```

Inspect the normalized object.
```{r step07-inspect-normalized-object}
str(preprocessNoob_results)
class(preprocessNoob_results)
preprocessNoob_results
```

Extract normalized beta and M values.
```{r step07-extract-normalized-beta-m}
beta_preprocessNoob <- getBeta(preprocessNoob_results)
M_preprocessNoob <- getM(preprocessNoob_results)

head(beta_preprocessNoob)
dim(beta_preprocessNoob)
```

Save normalized objects for reproducibility.
```{r step07-save-normalized-results}
saveRDS(preprocessNoob_results, file=file.path(rdsDir, "preprocessNoob_results.rds"))
saveRDS(beta_preprocessNoob, file=file.path(rdsDir, "beta_preprocessNoob.rds"))
saveRDS(M_preprocessNoob, file=file.path(rdsDir, "M_preprocessNoob.rds"))
```

Split normalized beta values by probe chemistry.
```{r step07-split-normalized-beta-by-chemistry}
beta_preprocessNoob_I <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfI$Name,]
beta_preprocessNoob_II <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfII$Name,]
```

Calculate mean normalized beta values by probe chemistry.
```{r step07-mean-normalized-beta-by-chemistry}
mean_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I, 1, mean, na.rm=T)
mean_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II, 1, mean, na.rm=T)
```

Estimate density of normalized beta means.
```{r step07-density-normalized-beta-by-chemistry}
d_mean_of_beta_preprocessNoob_I <- density(mean_of_beta_preprocessNoob_I, na.rm=T)
d_mean_of_beta_preprocessNoob_II <- density(mean_of_beta_preprocessNoob_II, na.rm=T)
```

Calculate and estimate density of normalized beta standard deviations.
```{r step07-normalized-beta-sd-by-chemistry}
sd_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I, 1, sd, na.rm=T)
sd_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II, 1, sd, na.rm=T)

d_sd_of_beta_preprocessNoob_I <- density(sd_of_beta_preprocessNoob_I, na.rm=T)
d_sd_of_beta_preprocessNoob_II <- density(sd_of_beta_preprocessNoob_II, na.rm=T)
```

Define colors for the raw and normalized beta boxplots.
```{r step07-define-boxplot-colors}
pheno <- data.frame(pData(RGset))
pheno$Group

boxplot_col <- c(CTRL = "blue", DIS = "red")[as.character(pheno$Group)]
boxplot_col
```

Compare raw and normalized beta values using plots.
```{r step07-compare-beta-chemistry, message=FALSE, warning=FALSE}
# Plot raw and normalized beta-value comparisons
par(mfrow = c(2, 3), mar = c(4.5, 4.5, 3, 1), oma = c(1, 1, 1, 5))

plot(d_mean_of_beta_I, col = "blue", main = "raw beta", xlim = c(0, 1), ylim = c(0, 5))
lines(d_mean_of_beta_II, col="red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

plot(d_sd_of_beta_I, col="blue", main="raw sd", xlim=c(0,0.6), ylim=c(0,60))
lines(d_sd_of_beta_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

boxplot(beta, ylim = c(0, 1), col = boxplot_col, main = "raw beta", names=FALSE)
axis( 1, at = seq_len(ncol(beta)), labels = seq_len(ncol(beta)), cex.axis = 0.8)
legend( "right", inset = c(-0.4, 0), legend = c("CTRL", "DIS"), fill = c("blue", "red"), bty = "n", cex = 0.8, xpd = NA)

plot(d_mean_of_beta_preprocessNoob_I, col="blue", main="preprocessNoob beta", xlim=c(0,1), ylim=c(0,5))
lines(d_mean_of_beta_preprocessNoob_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

plot(d_sd_of_beta_preprocessNoob_I, col="blue", main="preprocessNoob sd", xlim=c(0,0.6), ylim=c(0,60))
lines(d_sd_of_beta_preprocessNoob_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

boxplot(beta_preprocessNoob, ylim = c(0, 1), col = boxplot_col, main = "preprocessNoob beta", names=FALSE)
axis( 1, at = seq_len(ncol(beta_preprocessNoob)), labels = seq_len(ncol(beta_preprocessNoob)), cex.axis = 0.8)
legend( "right", inset = c(-0.4, 0), legend = c("CTRL", "DIS"), fill = c("blue", "red"), bty = "n", cex = 0.8, xpd = NA)
```

**Interpretation:** In the density plots, blue lines represent Type I probes and red lines represent Type II probes, while in the boxplots blue boxes represent CTRL samples and red boxes represent DIS samples. After normalization with `preprocessNoob`, the global beta-value distributions remained within the expected 0–1 range and preserved the typical bimodal methylation pattern. Type I and Type II probes still showed different density profiles, with Type I probes showing stronger peaks toward the unmethylated and highly methylated extremes, whereas Type II probes were shifted toward more intermediate beta values. The standard deviation density plots showed higher variability for Type II probes than for Type I probes.
After normalization, the SD distribution changed slightly, suggesting that preprocessNoob modified part of the technical variability while preserving the overall beta-value structure.

The preprocessNoob method is highly appropriate for this specific dataset. Because our cohort is very small (only 8 samples), applying aggressive between-array normalizations (like preprocessQuantile, which forces all sample distributions to be identical) could over-correct the data and erase genuine biological differences. preprocessNoob is a within-array method that successfully performs background correction and dye-bias equalization without relying on large sample sizes. Furthermore, when comparing the boxplots colored by group (CTRL vs DIS) before and after normalization, the overall medians and interquartile ranges remain largely consistent. The overall distributions between the two groups do not show extreme, artifactual shifts, confirming that the samples are comparable and the normalization successfully preserved the core biological structure.

# Step 08. PCA on Normalized Beta Values 
PCA evaluates global structure and potential confounders
```{r load-data, message=FALSE, warning=FALSE}
#load normalized beta values 
beta_preprocessNoob <- readRDS("results/rds/beta_preprocessNoob.rds")

#load RGset object 
load("results/rds/RGset_Report.RData")

# Extract sample information
sample_info <- data.frame(pData(RGset))
```

## Perform PCA 
```{r pca-analysis}
#Run PCA on normalized beta values
pca_results <- prcomp(t(beta_preprocessNoob), scale.=TRUE)
#Variance explained
pca_var <- pca_results$sdev^2
pca_var_explained <- pca_var / sum(pca_var)
```

## Scree Plot
**Interpretation**: PC1 dominates with 42.7% of variance, followed by PC2 at 20.4%. Together they capture ~63% of total variance, which is reasonable for an 8-sample methylation dataset. The steep drop after PC2 and gradual tapering afterward suggests the major sources of biological/technical variation are captured in the first two components.


```{r scree-plot, fig.width=8, fig.height=6}
bp <- barplot(
  pca_var_explained[1:8],
  names.arg = paste0("PC", 1:8),
  ylim = c(0, 0.5),
  main = "Variance Explained by Principal Components",
  xlab = "Principal Components",
  ylab = "Proportion of Variance Explained"
)
text(
  bp,
  y = pca_var_explained[1:8],
  labels = paste0(round(pca_var_explained[1:8] * 100, 1), "%"),
  pos = 3,
  cex = 0.8
)
```

## PCA Colored by Disease Group

```{r pca-group, fig.width=8, fig.height=6}

xlim_pca <- range(pca_results$x[, "PC1"]) + c(-50, 100)
ylim_pca <- range(pca_results$x[, "PC2"]) + c(-50, 50)

plot(
  pca_results$x[,1],
  pca_results$x[,2],
  col = as.factor(sample_info$Group),
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Group"
)

text(
  pca_results$x[,1],
  pca_results$x[,2],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(as.factor(sample_info$Group)),
  col = seq_along(levels(as.factor(sample_info$Group))),
  pch = 17
)
```

**Interpretation**: When coloring the PCA by the biological group (CTRL vs DIS), the samples do not form distinct clusters and are intermixed across both primary principal components. This indicates that the disease status is not the primary driver of global methylation variance in this dataset. The relatively subtle epigenetic signature of the disease is being overshadowed by other factors capturing much larger chunks of the variance.

## PCA Colored by Sex

```{r pca-sex, fig.width=8, fig.height=6}
sample_info$Sex <- as.factor(sample_info$Sex)

plot(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  col = sample_info$Sex,
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Sex"
)

text(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(sample_info$Sex),
  col = seq_along(levels(sample_info$Sex)),
  pch = 17
)
```

**Interpretation:** DNA methylation is profoundly influenced by biological sex, primarily due to X-chromosome inactivation in females and sex-specific global methylation patterns. Looking at our plot, the samples perfectly divide along the second principal component (PC2, the y-axis) according to their sex. Because our samples separate flawlessly by sex but not by group, it confirms that biological sex is a massive driver of variance in this dataset, greatly overshadowing the disease-related epigenetic signals.

## PCA Colored by Batch (Slide)

```{r pca-slide, fig.width=8, fig.height=6}
sample_info$Slide <- as.factor(sample_info$Slide)

plot(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  col = sample_info$Slide,
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Sentrix_ID"
)

text(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(sample_info$Slide),
  col = seq_along(levels(sample_info$Slide)),
  pch = 17
)
```

**Interpretation:** The array slide (represented by the Sentrix_ID column) is a purely technical variable. We check the data for technical "batch effects" by coloring the PCA by Slide. As seen in our plot, the samples distinctly split into two clusters along the first principal component (PC1, the x-axis) based strictly on which slide they were processed on. Because PC1 represents the largest source of variance in the entire dataset, this strongly indicates that a severe technical artifact (batch effect) is still present. This introduces that while preprocessNoob is excellent at correcting background fluorescence and dye bias, it does not completely eliminate slide-to-slide batch effects. In a full-scale study, an additional batch-correction algorithm (such as ComBat) would be required before proceeding to differential methylation testing.

# Step 09. Differential Methylation Analysis with t-test

Differential methylation analysis identifies CpG-specific changes associated with disease status. Here, probe-wise t-tests compare normalized beta values between CTRL and DIS samples.

## Load Data

```{r load-dm-data}
load("results/rds/RGset_Report.RData")

beta_preprocessNoob <- readRDS(
  "results/rds/beta_preprocessNoob.rds"
)

sample_info <- data.frame(
  pData(RGset)
)

beta_matrix <- beta_preprocessNoob
sample_data <- sample_info
```

## Define Sample Groups: control and disease

```{r define-groups}
control_samples <- sample_data$Group == "CTRL"
disease_samples <- sample_data$Group == "DIS"
```

## Calculate Mean Beta Values

```{r mean-beta}
mean_ctrl <- rowMeans(
  beta_matrix[, control_samples]
)

mean_dis <- rowMeans(
  beta_matrix[, disease_samples]
)

delta_beta <- mean_dis - mean_ctrl
```

## Perform Probe-wise t-Tests

```{r t-tests}
p_values <- apply(
  beta_matrix,
  1,
  function(probe_values) {
    t.test(
      probe_values[control_samples],
      probe_values[disease_samples]
    )$p.value
  }
)
```

## Create Results Table

```{r results-table}
t_test_results <- data.frame(
  ProbeID = rownames(beta_matrix),
  Mean_CTRL = mean_ctrl,
  Mean_DIS = mean_dis,
  Delta_Beta = delta_beta,
  P_Value = p_values
)

head(t_test_results)
```

**Interpretation:** The probe-wise t-test produced results for 485,512 CpG probes. Using nominal p-values, 11,928 probes have `P_Value <= 0.05`; however, these results should be considered preliminary because multiple testing correction has not yet been applied. Since hundreds of thousands of CpG sites are tested simultaneously, nominally significant probes may include false positives, so Bonferroni and Benjamini-Hochberg correction will be needed in Step 10.

## Save Results

```{r save-results}
write.csv(
  t_test_results,
  "results/tables/step09_t_test_results.csv",
  row.names = FALSE
)
```

# Step 10. Multiple Testing Correction

In Step 09, we performed independent t-tests across approximately 450,000 CpG probes to identify differentially methylated probes between the Control (CTRL) and Disease (DIS) groups. Using a nominal p-value threshold of 0.05 without adjustment would lead to tens of thousands of false positives due to the sheer volume of tests being conducted. To control for this, we applied two multiple testing correction methods: Bonferroni (which strictly controls the Family-Wise Error Rate) and Benjamini-Hochberg (which controls the False Discovery Rate, or FDR).

```{r step10-multiple-testing, message=FALSE, warning=FALSE}
# Load the t-test results from Step 09
t_test_results <- read.csv("results/tables/step09_t_test_results.csv")

# 1. Apply Bonferroni correction
t_test_results$P_Bonferroni <- p.adjust(t_test_results$P_Value, method = "bonferroni")

# 2. Apply Benjamini-Hochberg (BH / FDR) correction
t_test_results$P_FDR <- p.adjust(t_test_results$P_Value, method = "BH")

# Define the significance threshold assigned in the pipeline
alpha <- 0.05

# Count the number of significant probes for each method
sig_nominal <- sum(t_test_results$P_Value <= alpha, na.rm = TRUE)
sig_bonferroni <- sum(t_test_results$P_Bonferroni <= alpha, na.rm = TRUE)
sig_fdr <- sum(t_test_results$P_FDR <= alpha, na.rm = TRUE)

# Create a summary table to answer the pipeline questions
correction_summary <- data.frame(
  Method = c("Nominal P-Value", "Bonferroni Correction", "BH (FDR) Correction"),
  Significant_Probes = c(sig_nominal, sig_bonferroni, sig_fdr)
)

print(correction_summary)

# Save the updated dataframe containing adjusted p-values for Steps 11 and 12
write.csv(t_test_results, "results/tables/step10_t_test_results_adjusted.csv", row.names = FALSE)
```

### **Interpretation and Discussion (Step 10)**
As demonstrated by the output table, we identify thousands of probes with a nominal p-value below 0.05 before applying any correction. However, after applying the rigorous Bonferroni and Benjamini-Hochberg corrections, zero probes remain significant at the 0.05 threshold. 

This stark drop-off highlights a crucial concept in high-dimensional data analysis: while many probes appear significant initially, most of these results are likely due to chance when running 450,000 simultaneous tests. 

Furthermore, the lack of genome-wide significant results post-correction could be attributed to a few biological and technical factors:
* **Subtle Biological Effect:** The epigenetic differences between the CTRL and DIS groups might be highly localized or have very small effect sizes, making them hard to detect without a massive sample size.
* **Normalization Limitations:** As we used `preprocessNoob`—which is considered a "light" background correction method—uncorrected technical noise or batch effects might be swamping the true biological signal. A more robust normalization approach, such as `preprocessQuantile` (which forces distributions to be identical across arrays), might yield more statistically significant probes post-correction.

---

# Step 11. Volcano and Manhattan Plots

Because our stringent FDR correction eliminated our statistically significant probes, we must rely on nominal p-values combined with the biological magnitude of change (Delta Beta) to identify "suggestive" loci for further exploratory analysis. We visualize these results using Volcano and Manhattan plots.

### Volcano Plot

The Volcano plot contrasts the biological effect size (x-axis, representing the Delta Beta or methylation difference) against statistical significance (y-axis, representing the negative log10 of the nominal p-value). We apply a threshold of a 10% methylation difference (|Delta Beta| > 0.1) alongside a nominal p-value < 0.05 to highlight biologically interesting probes.

```{r step11-volcano-plot, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
# Plot the Volcano Plot
plot(
  x = t_test_results$Delta_Beta,
  y = -log10(t_test_results$P_Value),
  pch = 16,
  cex = 0.5,
  col = "darkgray",
  xlab = "Delta Beta (DIS - CTRL)",
  ylab = "-log10(Nominal P-value)",
  main = "Volcano Plot of DNA Methylation",
  xlim = c(-max(abs(t_test_results$Delta_Beta), na.rm=TRUE), 
            max(abs(t_test_results$Delta_Beta), na.rm=TRUE))
)

# Highlight significant targets (Nominal P <= 0.05 and |Delta Beta| > 0.1)
hyper <- which(t_test_results$Delta_Beta > 0.1 & t_test_results$P_Value <= 0.05)
hypo <- which(t_test_results$Delta_Beta < -0.1 & t_test_results$P_Value <= 0.05)

points(t_test_results$Delta_Beta[hyper], -log10(t_test_results$P_Value)[hyper], col = "red", pch = 16, cex = 0.6)
points(t_test_results$Delta_Beta[hypo], -log10(t_test_results$P_Value)[hypo], col = "blue", pch = 16, cex = 0.6)

# Add horizontal and vertical lines for thresholds
abline(h = -log10(0.05), col = "black", lty = 2)
abline(v = c(-0.1, 0.1), col = "black", lty = 2)

legend("topright", 
       legend = c("Hypermethylated", "Hypomethylated"),
       col = c("red", "blue", "darkgray"), 
       pch = 16, cex = 0.8, bty="n")
```

**Interpretation:** The Volcano plot allows us to immediately spot probes that undergo severe hypermethylation (red dots, top right) or hypomethylation (blue dots, top left) in the disease state compared to the controls. While these may not survive FDR correction, loci showing a high Delta Beta are primary candidates for targeted validation studies, as large methylation swings are often biologically impactful regardless of array-wide statistical penalties.

### Manhattan Plot

The Manhattan plot maps our nominal p-values across the genomic coordinates (Chromosomes 1-22, X, and Y). This allows us to look for "hotspots"—specific genomic regions or chromosomes where differential methylation is clustered.

```{r step11-manhattan-plot, fig.width=10, fig.height=6, message=FALSE, warning=FALSE}
library(qqman)

# Load the normalized object to extract genomic annotations
preprocessNoob_results <- readRDS("results/rds/preprocessNoob_results.rds")
ann <- getAnnotation(preprocessNoob_results)

# Extract locations into a dataframe
locations <- data.frame(ProbeID = rownames(ann), CHR = ann$chr, MAPINFO = ann$pos)

# Merge coordinates with our t-test results
manhattan_data <- merge(t_test_results, locations, by = "ProbeID")

# Clean CHR column to be numeric for qqman (1-22, 23 for X, 24 for Y)
manhattan_data$CHR <- gsub("chr", "", manhattan_data$CHR)
manhattan_data$CHR[manhattan_data$CHR == "X"] <- 23
manhattan_data$CHR[manhattan_data$CHR == "Y"] <- 24
manhattan_data$CHR <- as.numeric(manhattan_data$CHR)

manhattan_data <- manhattan_data[!is.na(manhattan_data$CHR) & !is.na(manhattan_data$P_Value), ]

# Plot the Manhattan Plot
manhattan(
  manhattan_data,
  chr = "CHR",
  bp = "MAPINFO",
  snp = "ProbeID",
  p = "P_Value",
  col = c("gray50", "gray80"),
  suggestiveline = FALSE, 
  genomewideline = FALSE,
  main = "Manhattan Plot of Epigenome-Wide Association",
  ylab = "-log10(Nominal P-value)"
)
```

**Interpretation:** If differential methylation is randomly scattered due to technical noise, the Manhattan plot will look like a uniform band across all chromosomes. On the contrary, if the disease state strongly affects specific genes or genomic regions, therefore, we expect to see distinct vertical "towers" of highly significant points rising above the background.

---

# Step 12. Heatmap of the Top 100 CpGs

To visually assess whether the disease signature is distinct from the control signature, we isolated the top 100 most significant probes (based on the lowest nominal p-values) and generated a heatmap. We applied unsupervised hierarchical clustering to both the probes (rows) and the samples (columns).

```{r step12-heatmap, fig.width=8, fig.height=8, message=FALSE, warning=FALSE}
library(gplots)

# Load necessary data
beta_preprocessNoob <- readRDS("results/rds/beta_preprocessNoob.rds")
load("results/rds/RGset_Report.RData")
pheno <- data.frame(pData(RGset))

# Sort the results by the lowest nominal p-value to extract the top 100 probes
top_100_results <- t_test_results[order(t_test_results$P_Value), ][1:100, ]
top_100_probes <- top_100_results$ProbeID

# Subset the normalized beta matrix to include only these 100 probes
beta_top100 <- beta_preprocessNoob[top_100_probes, ]

# Define column colors based on sample group (CTRL vs DIS)
group_colors <- ifelse(pheno$Group == "CTRL", "blue", "red")

# Generate the heatmap using complete linkage
heatmap.2(
  as.matrix(beta_top100),
  main = "Top 100 Differentially Methylated CpGs",
  trace = "none",              
  col = bluered(100),          
  scale = "none",              
  dendrogram = "both",         
  ColSideColors = group_colors,
  margins = c(10, 5),          
  cexCol = 0.8,
  labRow = FALSE               
)

# Add a legend for the column side colors
legend("topright", 
       legend = levels(as.factor(pheno$Group)), 
       fill = c("blue", "red"), 
       border = FALSE, 
       bty = "n", 
       cex = 0.8)
```

### **Interpretation and Final Conclusion (Step 12)**
The heatmap provides a crucial sanity check for our data structure. Look closely at the dendrogram (the clustering tree) at the top of the columns:
* **If the clustering is successful:** The tree will strictly separate the blue markers (CTRL) from the red markers (DIS) into two distinct branches. This means that, despite the lack of genome-wide strict FDR significance, these top 100 loci possess a strong, coordinated epigenetic signature capable of differentiating a diseased sample from a healthy one. 
* **If the clustering is mixed:** If CTRL and DIS samples intermingle on the dendrogram branches, it implies that the top 100 probes are capturing individual patient heterogeneity, batch effects, or random noise rather than a unifying disease mechanism. 