---
title: "Comprehensive Analysis of DNA Methylation Using the Illumina HumanMethylation450 BeadChip Platform"
subtitle: "DNA & RNA Dynamics final project, Team 1. Repository: [Parganini/DRD-Final-Project-Team1](https://github.com/Parganini/DRD-Final-Project-Team1)"
author: "Fatemeh Bayat, Tamime Eter, Seyed Parsa Golshani, Amirhossein Heidari, Mahmoud Ahmed Mohamed Mohamed Ibrahim, Hector Eduardo Parga Najera, Bita Rahmani, Shayan Zanjaniha"
date: "2026-06"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,
                      fig.align="center", dpi=300, fig.retina=2)
knitr::opts_knit$set(root.dir=normalizePath("."))
```

# Introduction


Team 1 assignments:

- Address for the fluorescence check: `42796479`
- Detection p-value threshold: `0.05`
- Normalization method: `preprocessNoob`
- Differential methylation test: `t-test`

# Step 00. Basic Preparation

Clean the R environment

```{r clean-environment}
rm(list=ls())
```

Load the packages required for this analysis

```{r load-packages}
library(minfi)
library(qqman)
library(gplots)
library(IlluminaHumanMethylation450kmanifest)
```

Define the folders used by the project

```{r define-folders}
baseDir <- "data/raw"
rdsDir <- "results/rds"
tablesDir <- "results/tables"
figuresDir <- "results/figures"
```

# Step 01. Data Import and RGChannelSet Creation

```{r step01-load-data}
list.files("./data/raw/")
SampleSheet <- read.csv("./data/raw/SampleSheet_Report_II.csv", header = TRUE)
SampleSheet

targets <- read.metharray.sheet("./data/raw")
targets

RGset <- read.metharray.exp(targets = targets)
save(RGset, file = file.path(rdsDir, "RGset_Report.RData"))

RGset
str(RGset)
```

The `SampleSheet` contains the metadata for our 8 samples, including group (CTRL/DIS), sex, and Sentrix ID/Position. `read.metharray.sheet()` reads this sheet and automatically attaches the file paths needed to locate the corresponding idat files. `read.metharray.exp()` then loads the raw Red and Green fluorescence intensities for all samples into an `RGChannelSet` object called `RGset`.

# Step 02. Red and Green Fluorescence Channels

```{r step02-extract-channels}
Red <- data.frame(getRed(RGset))
dim(Red)
head(Red)

Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)
```

`getRed()` and `getGreen()` extract the raw fluorescence intensity matrices from `RGset`, one row per probe address and one column per sample.

# Step 03. Team 1 Address Check

```{r step03-address-lookup}
TARGET_ADDRESS <- "42796479"

getManifest(RGset)
ProbeInfo_I  <- data.frame(getProbeInfo(RGset, type = "I"))
ProbeInfo_II <- data.frame(getProbeInfo(RGset, type = "II"))

ProbeInfo_I[ProbeInfo_I$AddressA == TARGET_ADDRESS, ]
ProbeInfo_I[ProbeInfo_I$AddressB == TARGET_ADDRESS, ]
ProbeInfo_II[ProbeInfo_II$AddressA == TARGET_ADDRESS, ]

Red_fluorescences   <- Red[rownames(Red) == TARGET_ADDRESS, ]
Green_fluorescences <- Green[rownames(Green) == TARGET_ADDRESS, ]

df_summary_probes <- data.frame(
  Sample             = colnames(Red_fluorescences),
  Red_Fluorescence   = as.numeric(Red_fluorescences[1, ]),
  Green_Fluorescence = as.numeric(Green_fluorescences[1, ]),
  Type               = "I",
  Color              = "Red"
)

print(df_summary_probes)
```

Address `42796479` corresponds to the **AddressA (unmethylated)** position of probe **cg10602367**, a **Type I, Red-channel** probe. Type I probes use a single color channel, so the Green fluorescence values for this address are near background levels across all samples, while Red fluorescence varies between samples — reflecting biological differences in signal intensity at this CpG site.

# Step 04. MethylSet Creation

```{r step04-mset-raw}
MSet.raw <- preprocessRaw(RGset)
MSet.raw
save(MSet.raw, file = file.path(rdsDir, "MSet_raw.RData"))

Meth <- as.matrix(getMeth(MSet.raw))
str(Meth)
head(Meth)

Unmeth <- as.matrix(getUnmeth(MSet.raw))
str(Unmeth)
head(Unmeth)

Meth[rownames(Meth) == "cg10602367",]
Unmeth[rownames(Unmeth) == "cg10602367",]
```

`preprocessRaw()` converts the raw Red/Green intensities in `RGset` into a `MethylSet`, splitting the signal into **Methylated** and **Unmethylated** intensities for each probe. For our probe `cg10602367`, the Unmethylated signal is substantially higher than the Methylated signal across all 8 samples, consistent with the Red-channel Type I design identified in Step 3.

# Step 05. Quality Control

```{r step05-qc-plot}
qc <- getQC(MSet.raw)
qc
plotQC(qc)
```

The QC plot shows the median methylated (mMed) and unmethylated (uMed) signal intensities per sample. Samples with higher values in both signals indicate higher quality. Most samples show median values in the 8.6–11.2 range (log2 scale); a few samples (e.g. GSM5319603, GSM5319609) show comparatively lower values, suggesting somewhat lower signal quality.

```{r step05-negative-controls}
NegativeControl <- data.frame(getProbeInfo(RGset, type = "Control"))
table(NegativeControl$Type)
head(NegativeControl)
controlStripPlot(RGset, controls = "NEGATIVE")
```

The negative control probes show signal intensities below log2(10) in both channels across all samples, indicating acceptable background noise levels.

```{r step05-detection-pvalues}
Detection_pvalue <- detectionP(RGset)
save(Detection_pvalue, file = file.path(rdsDir, "Detection_pvalue.RData"))

Over_Threshold <- Detection_pvalue > 0.05
table(Over_Threshold)
sum(Over_Threshold)

Total_Failed_Positions <- sum(Over_Threshold)
Total_Detection_Tests <- length(Over_Threshold)
Total_Failed_Percentage <- Total_Failed_Positions / Total_Detection_Tests * 100

failed_probe_message <- sprintf(
  "There are %s failed detection tests with a detection p-value higher than 0.05",
  Total_Failed_Positions
)

Over_Threshold_Per_Sample <- colSums(Over_Threshold)
Over_Threshold_Per_Sample

Over_Threshold_Fraction_Per_Sample <- colMeans(Over_Threshold)
Over_Threshold_Fraction_Per_Sample

failed <- data.frame(
  Sample = names(Over_Threshold_Per_Sample),
  n_Failed_Positions = Over_Threshold_Per_Sample,
  Fraction_Failed_Positions = Over_Threshold_Fraction_Per_Sample,
  Percentage_Failed_Positions = Over_Threshold_Fraction_Per_Sample * 100
)
print(failed)
print(failed_probe_message)
```

Using Team 1's detection p-value threshold of 0.05, there are `r Total_Failed_Positions` failed detection tests across all samples, corresponding to `r round(Total_Failed_Percentage, 2)`% of all probe-sample detection tests. Per sample, the number of failed positions ranges from `r min(Over_Threshold_Per_Sample)` to `r max(Over_Threshold_Per_Sample)`, corresponding to `r round(min(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% to `r round(max(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% of positions. Since all samples show a low percentage of failed positions, the overall detection quality is good and all samples can be kept for the next steps.

# Step 06. Raw Beta and M Values



# Step 07. Normalization with preprocessNoob



# Step 08. PCA on Normalized Beta Values



# Step 09. Differential Methylation Analysis with t-test



# Step 10. Multiple Testing Correction



# Step 11. Volcano and Manhattan Plots



# Step 12. Heatmap of the Top 100 CpGs



# Conclusions


