---
title: "Comprehensive Analysis of DNA Methylation Using the Illumina HumanMethylation450 BeadChip Platform"
subtitle: "DNA & RNA Dynamics final project, Team 1. Repository: [Parganini/DRD-Final-Project-Team1](https://github.com/Parganini/DRD-Final-Project-Team1)"
author: "Fatemeh Bayat, Tamime Eter, Seyed Parsa Golshani, Amirhossein Heidari, Mahmoud Ahmed Mohamed Mohamed Ibrahim, Hector Eduardo Parga Najera, Bita Rahmani, Shayan Zanjaniha"
date: "2026-06"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,
                      fig.align="center", dpi=300, fig.retina=2)
knitr::opts_knit$set(root.dir=normalizePath("."))
```

# Introduction


Team 1 assignments:

- Address for the fluorescence check: `42796479`
- Detection p-value threshold: `0.05`
- Normalization method: `preprocessNoob`
- Differential methylation test: `t-test`

# Step 00. Basic Preparation

Clean the R environment

```{r clean-environment}
rm(list=ls())
```

Load the packages required for this analysis

```{r load-packages}
library(minfi)
library(qqman)
library(gplots)
library(IlluminaHumanMethylation450kmanifest)
```

Define the folders used by the project

```{r define-folders}
baseDir <- "data/raw"
rdsDir <- "results/rds"
tablesDir <- "results/tables"
figuresDir <- "results/figures"
```

# Step 01. Data Import and RGChannelSet Creation

```{r step01-load-data}
list.files("./data/raw/")
SampleSheet <- read.csv("./data/raw/SampleSheet_Report_II.csv", header = TRUE)
SampleSheet

targets <- read.metharray.sheet("./data/raw")
targets

RGset <- read.metharray.exp(targets = targets)
save(RGset, file = file.path(rdsDir, "RGset_Report.RData"))

RGset
str(RGset)
```

The `SampleSheet` contains the metadata for our 8 samples, including group (CTRL/DIS), sex, and Sentrix ID/Position. `read.metharray.sheet()` reads this sheet and automatically attaches the file paths needed to locate the corresponding idat files. `read.metharray.exp()` then loads the raw Red and Green fluorescence intensities for all samples into an `RGChannelSet` object called `RGset`.

# Step 02. Red and Green Fluorescence Channels

```{r step02-extract-channels}
Red <- data.frame(getRed(RGset))
dim(Red)
head(Red)

Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)
```

`getRed()` and `getGreen()` extract the raw fluorescence intensity matrices from `RGset`, one row per probe address and one column per sample.

# Step 03. Team 1 Address Check

```{r step03-address-lookup}
TARGET_ADDRESS <- "42796479"

getManifest(RGset)
ProbeInfo_I  <- data.frame(getProbeInfo(RGset, type = "I"))
ProbeInfo_II <- data.frame(getProbeInfo(RGset, type = "II"))

ProbeInfo_I[ProbeInfo_I$AddressA == TARGET_ADDRESS, ]
ProbeInfo_I[ProbeInfo_I$AddressB == TARGET_ADDRESS, ]
ProbeInfo_II[ProbeInfo_II$AddressA == TARGET_ADDRESS, ]

Red_fluorescences   <- Red[rownames(Red) == TARGET_ADDRESS, ]
Green_fluorescences <- Green[rownames(Green) == TARGET_ADDRESS, ]

df_summary_probes <- data.frame(
  Sample             = colnames(Red_fluorescences),
  Red_Fluorescence   = as.numeric(Red_fluorescences[1, ]),
  Green_Fluorescence = as.numeric(Green_fluorescences[1, ]),
  Type               = "I",
  Color              = "Red"
)

print(df_summary_probes)
```

Address `42796479` corresponds to the **AddressA (unmethylated)** position of probe **cg10602367**, a **Type I, Red-channel** probe. Type I probes use a single color channel, so the Green fluorescence values for this address are near background levels across all samples, while Red fluorescence varies between samples — reflecting biological differences in signal intensity at this CpG site.

# Step 04. MethylSet Creation

```{r step04-mset-raw}
MSet.raw <- preprocessRaw(RGset)
MSet.raw
save(MSet.raw, file = file.path(rdsDir, "MSet_raw.RData"))

Meth <- as.matrix(getMeth(MSet.raw))
str(Meth)
head(Meth)

Unmeth <- as.matrix(getUnmeth(MSet.raw))
str(Unmeth)
head(Unmeth)

Meth[rownames(Meth) == "cg10602367",]
Unmeth[rownames(Unmeth) == "cg10602367",]
```

`preprocessRaw()` converts the raw Red/Green intensities in `RGset` into a `MethylSet`, splitting the signal into **Methylated** and **Unmethylated** intensities for each probe. For our probe `cg10602367`, the Unmethylated signal is substantially higher than the Methylated signal across all 8 samples, consistent with the Red-channel Type I design identified in Step 3.

# Step 05. Quality Control

```{r step05-qc-plot}
qc <- getQC(MSet.raw)
qc
plotQC(qc)
```

**Interpretation:** The QC plot shows the median methylated (mMed) and unmethylated (uMed) signal intensities per sample. Samples with higher values in both signals indicate higher quality. Most samples show median values in the 8.6–11.2 range (log2 scale); a few samples (e.g. GSM5319603, GSM5319609) show comparatively lower values, suggesting somewhat lower signal quality.

```{r step05-negative-controls}
NegativeControl <- data.frame(getProbeInfo(RGset, type = "Control"))
table(NegativeControl$Type)
head(NegativeControl)
controlStripPlot(RGset, controls = "NEGATIVE")
```

The negative control probes show signal intensities below log2(10) in both channels across all samples, indicating acceptable background noise levels.

```{r step05-detection-pvalues}
Detection_pvalue <- detectionP(RGset)
save(Detection_pvalue, file = file.path(rdsDir, "Detection_pvalue.RData"))

Over_Threshold <- Detection_pvalue > 0.05
table(Over_Threshold)
sum(Over_Threshold)

Total_Failed_Positions <- sum(Over_Threshold)
Total_Detection_Tests <- length(Over_Threshold)
Total_Failed_Percentage <- Total_Failed_Positions / Total_Detection_Tests * 100

failed_probe_message <- sprintf(
  "There are %s failed detection tests with a detection p-value higher than 0.05",
  Total_Failed_Positions
)

Over_Threshold_Per_Sample <- colSums(Over_Threshold)
Over_Threshold_Per_Sample

Over_Threshold_Fraction_Per_Sample <- colMeans(Over_Threshold)
Over_Threshold_Fraction_Per_Sample

failed <- data.frame(
  Sample = names(Over_Threshold_Per_Sample),
  n_Failed_Positions = Over_Threshold_Per_Sample,
  Fraction_Failed_Positions = Over_Threshold_Fraction_Per_Sample,
  Percentage_Failed_Positions = Over_Threshold_Fraction_Per_Sample * 100
)
print(failed)
print(failed_probe_message)
```

**Interpretation:** Using Team 1's detection p-value threshold of 0.05, there are `r Total_Failed_Positions` failed detection tests across all samples, corresponding to `r round(Total_Failed_Percentage, 2)`% of all probe-sample detection tests. Per sample, the number of failed positions ranges from `r min(Over_Threshold_Per_Sample)` to `r max(Over_Threshold_Per_Sample)`, corresponding to `r round(min(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% to `r round(max(Over_Threshold_Fraction_Per_Sample) * 100, 2)`% of positions. Since all samples show a low percentage of failed positions, the overall detection quality is good and all samples can be kept for the next steps.

# Step 06. Raw Beta and M Values

We retrieve beta and M value matrices from the raw MethylSet.

Extract raw beta values and inspect their dimensions and distribution.
```{r step06-beta-raw}
beta <- getBeta(MSet.raw)
class(beta)
dim(beta)
head(beta)
summary(beta)
```

Extract raw M values and inspect their dimensions and distribution.
```{r step06-m-raw}
M <- getM(MSet.raw)
dim(M)
head(M)
summary(M)
```

Save raw beta and M matrices.
```{r step06-save-beta-m}
saveRDS(beta, file=file.path(rdsDir, "beta_raw.rds"))
saveRDS(M, file=file.path(rdsDir, "M_raw.rds"))
```

Recover phenotype data to split samples by group.
```{r step06-phenotype-data}
pheno <- data.frame(pData(RGset))
str(pheno)
table(pheno$Group)
```

Split beta values into CTRL and DIS groups.
```{r step06-split-beta}
beta_CTRL <- beta[, pheno$Group=="CTRL"]
beta_DIS <- beta[, pheno$Group == "DIS"]

dim(beta_CTRL)
dim(beta_DIS)
```

Split M values into CTRL and DIS groups.
```{r step06-split-m}
M_CTRL <- M[, pheno$Group=="CTRL"]
M_DIS <- M[, pheno$Group=="DIS"]

dim(M_CTRL)
dim(M_DIS)
```

Calculate mean beta and M values per CpG within each group.
```{r step06-summary-beta-m}
mean_of_beta_CTRL <- apply(beta_CTRL, 1, mean, na.rm=TRUE)
mean_of_beta_DIS <- apply(beta_DIS, 1, mean, na.rm=TRUE)

mean_of_M_CTRL <- apply(M_CTRL, 1, mean, na.rm=TRUE)
mean_of_M_DIS <- apply(M_DIS, 1, mean, na.rm = TRUE)
```
Remove infinite M values before density estimation.
```{r step06-clean-infinite-m}
mean_of_M_CTRL <- mean_of_M_CTRL[is.finite(mean_of_M_CTRL)]
mean_of_M_DIS <- mean_of_M_DIS[is.finite(mean_of_M_DIS)]
```

Estimate density distributions for raw beta and M means.
```{r step06-density-estimation}
d_mean_of_beta_CTRL <- density(mean_of_beta_CTRL, na.rm=TRUE)
d_mean_of_beta_DIS <- density(mean_of_beta_DIS, na.rm=TRUE)

d_mean_of_M_CTRL <- density(mean_of_M_CTRL, na.rm=TRUE)
d_mean_of_M_DIS <- density(mean_of_M_DIS, na.rm=TRUE)
```

Plot raw beta density distributions by group.
```{r step06-plot-beta-density, message=FALSE, warning=FALSE}
plot(d_mean_of_beta_CTRL, col="blue",
     main="Raw beta values", xlab="Mean beta")

lines(d_mean_of_beta_DIS, col="red")

legend("topright", legend=c("CTRL", "DIS"),
       col=c("blue", "red"), lty=1, bty="n")
```

Plot raw M value density distributions by group.
```{r step06-plot-M-density, message=FALSE, warning=FALSE}
plot(d_mean_of_M_CTRL, col="blue",
     main="Raw M values", xlab="Mean M")

lines(d_mean_of_M_DIS, col="red")

legend("topright", legend=c("CTRL", "DIS"),
       col=c("blue", "red"), lty=1, bty="n")
```

**Interpretation:** The raw beta and M-value density curves for CTRL and DIS samples show broadly similar distributions. Small shifts between the two groups can be observed, but the overall methylation profiles are largely comparable at the raw-data level.

Save summary vectors used for the density plots.
```{r step06-save-summary-vectors}
saveRDS(mean_of_beta_CTRL,
        file=file.path(rdsDir, "mean_of_beta_raw_CTRL.rds"))

saveRDS(mean_of_beta_DIS,
        file=file.path(rdsDir, "mean_of_beta_raw_DIS.rds"))

saveRDS(mean_of_M_CTRL,
        file=file.path(rdsDir, "mean_of_M_raw_CTRL.rds"))

saveRDS(mean_of_M_DIS,
        file=file.path(rdsDir, "mean_of_M_raw_DIS.rds"))
```

# Step 07. Normalization with preprocessNoob

The assigned Team 1 normalization method was `preprocessNoob`. Raw and normalized beta values were compared by Type I and Type II probe chemistry. This comparison is important because Infinium Type I and Type II assays can show systematic differences: Type II beta values tend to be shifted toward intermediate methylation levels and can show different variability from Type I probes. Normalization is used to reduce technical variation before biological interpretation.

Open the raw RGChannelSet and raw beta matrix saved in previous steps.
```{r step07-load-raw-data}
load(file.path(rdsDir, "RGset_Report.RData"))
beta <- readRDS(file.path(rdsDir, "beta_raw.rds"))
```

Inspect the raw RGChannelSet and beta matrix.
```{r step07-inspect-raw-data}
RGset
head(beta)
dim(beta)
```

Retrieve Type I and Type II probe annotations from the manifest.
```{r step07-probe-annotations}
dfI <- data.frame(getProbeInfo(RGset, type="I"))
dfII <- data.frame(getProbeInfo(RGset, type="II"))

dim(dfI)
dim(dfII)
str(dfI)
str(dfII)
head(dfI$Name)
head(dfII$Name)
```

Split raw beta values by probe chemistry.
```{r step07-split-beta-by-chemistry}
beta_I <- beta[rownames(beta) %in% dfI$Name,]
beta_II <- beta[rownames(beta) %in% dfII$Name,]

dim(beta_I)
dim(beta_II)
```

Calculate mean raw beta by probe chemistry.
```{r step07-mean-beta-by-chemistry}
mean_of_beta_I <- apply(beta_I, 1, mean, na.rm=T)
mean_of_beta_II <- apply(beta_II, 1, mean, na.rm=T)
```

Estimate density of raw beta means by probe chemistry.
```{r step07-density-beta-by-chemistry}
d_mean_of_beta_I <- density(mean_of_beta_I, na.rm=T)
d_mean_of_beta_II <- density(mean_of_beta_II, na.rm=T)
```

Calculate and estimate density of raw beta standard deviations.
```{r step07-beta-sd-by-chemistry}
sd_of_beta_I <- apply(beta_I, 1, sd, na.rm=T)
sd_of_beta_II <- apply(beta_II, 1, sd, na.rm=T)

d_sd_of_beta_I <- density(sd_of_beta_I, na.rm=T)
d_sd_of_beta_II <- density(sd_of_beta_II, na.rm=T)
```
Normalize the raw RGChannelSet with preprocessNoob.
```{r step07-normalization}
preprocessNoob_results <- preprocessNoob(RGset)
```

Inspect the normalized object.
```{r step07-inspect-normalized-object}
str(preprocessNoob_results)
class(preprocessNoob_results)
preprocessNoob_results
```

Extract normalized beta and M values.
```{r step07-extract-normalized-beta-m}
beta_preprocessNoob <- getBeta(preprocessNoob_results)
M_preprocessNoob <- getM(preprocessNoob_results)

head(beta_preprocessNoob)
dim(beta_preprocessNoob)
```

Save normalized objects for reproducibility.
```{r step07-save-normalized-results}
saveRDS(preprocessNoob_results, file=file.path(rdsDir, "preprocessNoob_results.rds"))
saveRDS(beta_preprocessNoob, file=file.path(rdsDir, "beta_preprocessNoob.rds"))
saveRDS(M_preprocessNoob, file=file.path(rdsDir, "M_preprocessNoob.rds"))
```

Split normalized beta values by probe chemistry.
```{r step07-split-normalized-beta-by-chemistry}
beta_preprocessNoob_I <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfI$Name,]
beta_preprocessNoob_II <- beta_preprocessNoob[rownames(beta_preprocessNoob) %in% dfII$Name,]
```

Calculate mean normalized beta values by probe chemistry.
```{r step07-mean-normalized-beta-by-chemistry}
mean_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I, 1, mean, na.rm=T)
mean_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II, 1, mean, na.rm=T)
```

Estimate density of normalized beta means.
```{r step07-density-normalized-beta-by-chemistry}
d_mean_of_beta_preprocessNoob_I <- density(mean_of_beta_preprocessNoob_I, na.rm=T)
d_mean_of_beta_preprocessNoob_II <- density(mean_of_beta_preprocessNoob_II, na.rm=T)
```

Calculate and estimate density of normalized beta standard deviations.
```{r step07-normalized-beta-sd-by-chemistry}
sd_of_beta_preprocessNoob_I <- apply(beta_preprocessNoob_I, 1, sd, na.rm=T)
sd_of_beta_preprocessNoob_II <- apply(beta_preprocessNoob_II, 1, sd, na.rm=T)

d_sd_of_beta_preprocessNoob_I <- density(sd_of_beta_preprocessNoob_I, na.rm=T)
d_sd_of_beta_preprocessNoob_II <- density(sd_of_beta_preprocessNoob_II, na.rm=T)
```

Define colors for the raw and normalized beta boxplots.
```{r step07-define-boxplot-colors}
pheno <- data.frame(pData(RGset))
pheno$Group

boxplot_col <- c(CTRL = "blue", DIS = "red")[as.character(pheno$Group)]
boxplot_col
```

Compare raw and normalized beta values using plots.
```{r step07-compare-beta-chemistry, message=FALSE, warning=FALSE}
# Plot raw and normalized beta-value comparisons
par(mfrow = c(2, 3), mar = c(4.5, 4.5, 3, 1), oma = c(1, 1, 1, 5))

plot(d_mean_of_beta_I, col = "blue", main = "raw beta", xlim = c(0, 1), ylim = c(0, 5))
lines(d_mean_of_beta_II, col="red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

plot(d_sd_of_beta_I, col="blue", main="raw sd", xlim=c(0,0.6), ylim=c(0,60))
lines(d_sd_of_beta_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

boxplot(beta, ylim = c(0, 1), col = boxplot_col, main = "raw beta", names=FALSE)
axis( 1, at = seq_len(ncol(beta)), labels = seq_len(ncol(beta)), cex.axis = 0.8)
legend( "right", inset = c(-0.4, 0), legend = c("CTRL", "DIS"), fill = c("blue", "red"), bty = "n", cex = 0.8, xpd = NA)

plot(d_mean_of_beta_preprocessNoob_I, col="blue", main="preprocessNoob beta", xlim=c(0,1), ylim=c(0,5))
lines(d_mean_of_beta_preprocessNoob_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

plot(d_sd_of_beta_preprocessNoob_I, col="blue", main="preprocessNoob sd", xlim=c(0,0.6), ylim=c(0,60))
lines(d_sd_of_beta_preprocessNoob_II, col = "red")
legend("topright", legend = c("Type I", "Type II"), col = c("blue", "red"), lty = 1, bty = "n", cex = 0.8)

boxplot(beta_preprocessNoob, ylim = c(0, 1), col = boxplot_col, main = "preprocessNoob beta", names=FALSE)
axis( 1, at = seq_len(ncol(beta_preprocessNoob)), labels = seq_len(ncol(beta_preprocessNoob)), cex.axis = 0.8)
legend( "right", inset = c(-0.4, 0), legend = c("CTRL", "DIS"), fill = c("blue", "red"), bty = "n", cex = 0.8, xpd = NA)
```

**Interpretation:** In the density plots, blue lines represent Type I probes and red lines represent Type II probes, while in the boxplots blue boxes represent CTRL samples and red boxes represent DIS samples. After normalization with `preprocessNoob`, the global beta-value distributions remained within the expected 0–1 range and preserved the typical bimodal methylation pattern. Type I and Type II probes still showed different density profiles, with Type I probes showing stronger peaks toward the unmethylated and highly methylated extremes, whereas Type II probes were shifted toward more intermediate beta values. The standard deviation density plots showed higher variability for Type II probes than for Type I probes.
After normalization, the SD distribution changed slightly, suggesting that preprocessNoob modified part of the technical variability while preserving the overall beta-value structure.

# Step 08. PCA on Normalized Beta Values: PCA evaluates global structure and potential confounders
```{r load-data, message=FALSE, warning=FALSE}
#load normalized beta values 
beta_preprocessNoob <- readRDS("results/rds/beta_preprocessNoob.rds")

#load RGset object 
load("results/rds/RGset_Report.RData")

# Extract sample information
sample_info <- data.frame(pData(RGset))
```

## Perform PCA 
```{r pca-analysis}
#Run PCA on normalized beta values
pca_results <- prcomp(t(beta_preprocessNoob), scale.=TRUE)
#Variance explained
pca_var <- pca_results$sdev^2
pca_var_explained <- pca_var / sum(pca_var)
```

## Scree Plot
**Interpretation**: PC1 dominates with 42.7% of variance, followed by PC2 at 20.4%. Together they capture ~63% of total variance, which is reasonable for an 8-sample methylation dataset. The steep drop after PC2 and gradual tapering afterward suggests the major sources of biological/technical variation are captured in the first two components.


```{r scree-plot, fig.width=8, fig.height=6}
bp <- barplot(
  pca_var_explained[1:8],
  names.arg = paste0("PC", 1:8),
  ylim = c(0, 0.5),
  main = "Variance Explained by Principal Components",
  xlab = "Principal Components",
  ylab = "Proportion of Variance Explained"
)
text(
  bp,
  y = pca_var_explained[1:8],
  labels = paste0(round(pca_var_explained[1:8] * 100, 1), "%"),
  pos = 3,
  cex = 0.8
)
```

## PCA Colored by Disease Group

```{r pca-group, fig.width=8, fig.height=6}

xlim_pca <- range(pca_results$x[, "PC1"]) + c(-50, 100)
ylim_pca <- range(pca_results$x[, "PC2"]) + c(-50, 50)

plot(
  pca_results$x[,1],
  pca_results$x[,2],
  col = as.factor(sample_info$Group),
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Group"
)

text(
  pca_results$x[,1],
  pca_results$x[,2],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(as.factor(sample_info$Group)),
  col = seq_along(levels(as.factor(sample_info$Group))),
  pch = 17
)
```

**Interpretation**: The CTRL (black) and DIS (pink) samples do not separate cleanly along PC1 or PC2. Samples from both groups are intermixed across the plot space. This suggests that disease status alone is not the dominant driver of global methylation variance — other sources of variation (sex, batch) likely contribute more to the top PCs.

## PCA Colored by Sex

```{r pca-sex, fig.width=8, fig.height=6}
sample_info$Sex <- as.factor(sample_info$Sex)

plot(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  col = sample_info$Sex,
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Sex"
)

text(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(sample_info$Sex),
  col = seq_along(levels(sample_info$Sex)),
  pch = 17
)
```

**Interpretation**: Male samples (pink) and Female samples (black) show a partial separation, particularly along PC1. However, male and female samples do overlap in the central cluster, so sex is not entirely driving PC1.
Sex is often a major source of variance in methylation due to the X and Y chromosomes; if your differential methylation analysis doesn't account for this, sex could potentially act as a covariate in downstream differential methylation analysis to avoid false positives 

**Suggested Interpretation:** Female samples are shown in black and male samples are shown in pink/red. The PCA does not show a complete separation between male and female samples. Some samples are more distant from the central cluster, but sex alone does not fully explain PC1 and PC2. Since sex can influence methylation patterns, especially through sex chromosomes, it should be considered when interpreting downstream differential methylation results.

## PCA Colored by Batch (Slide)

```{r pca-slide, fig.width=8, fig.height=6}
sample_info$Slide <- as.factor(sample_info$Slide)

plot(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  col = sample_info$Slide,
  pch = 17,
  cex = 1.5,
  xlim = xlim_pca,
  ylim = ylim_pca,
  xlab = paste0(
    "PC1 (",
    round(pca_var_explained[1] * 100, 1),
    "%)"
  ),
  ylab = paste0(
    "PC2 (",
    round(pca_var_explained[2] * 100, 1),
    "%)"
  ),
  main = "PCA of Normalized Beta Values by Sentrix_ID"
)

text(
  pca_results$x[, "PC1"],
  pca_results$x[, "PC2"],
  labels = sample_info$SampleID,
  pos = 3,
  cex = 0.8
)

legend(
  "topright",
  legend = levels(sample_info$Slide),
  col = seq_along(levels(sample_info$Slide)),
  pch = 17
)
```

**Interpretation**: 
Five different slides (Sentrix IDs) are represented by different colors. There is no obvious clustering by slide, meaning batch effects from different arrays are not a major confounding factor in this dataset. Samples from different slides are spread across the PCA space without grouping, which is a positive quality control finding.

**Suggested Interpretation:** Five different slides/Sentrix IDs are represented. No single slide clearly explains all variation in PC1 and PC2, although samples from slide 3999547017 appear close to each other in the lower-left region. Because there are several slide IDs for only eight samples, batch effects cannot be ruled out completely, but slide does not appear to be the dominant source of variation in this PCA.

# Step 09. Differential Methylation Analysis with t-test

Differential methylation analysis identifies CpG-specific changes associated with disease status. Here, probe-wise t-tests compare normalized beta values between CTRL and DIS samples.

## Load Data

```{r load-dm-data}
load("results/rds/RGset_Report.RData")

beta_preprocessNoob <- readRDS(
  "results/rds/beta_preprocessNoob.rds"
)

sample_info <- data.frame(
  pData(RGset)
)

beta_matrix <- beta_preprocessNoob
sample_data <- sample_info
```

## Define Sample Groups: control and disease

```{r define-groups}
control_samples <- sample_data$Group == "CTRL"
disease_samples <- sample_data$Group == "DIS"
```

## Calculate Mean Beta Values

```{r mean-beta}
mean_ctrl <- rowMeans(
  beta_matrix[, control_samples]
)

mean_dis <- rowMeans(
  beta_matrix[, disease_samples]
)

delta_beta <- mean_dis - mean_ctrl
```

## Perform Probe-wise t-Tests

```{r t-tests}
p_values <- apply(
  beta_matrix,
  1,
  function(probe_values) {
    t.test(
      probe_values[control_samples],
      probe_values[disease_samples]
    )$p.value
  }
)
```

## Create Results Table

```{r results-table}
t_test_results <- data.frame(
  ProbeID = rownames(beta_matrix),
  Mean_CTRL = mean_ctrl,
  Mean_DIS = mean_dis,
  Delta_Beta = delta_beta,
  P_Value = p_values
)

head(t_test_results)
```

**Interpretation**: 
The t-test results show modest delta beta values (ranging roughly from -0.20 to +0.16) with no probe reaching statistical significance at this stage (all p-values > 0.05), which is expected before multiple testing correction and consistent with the small sample size of 8 individuals.
Although some probes exhibit biologically meaningful delta beta values (|Δβ| > 0.1) alongside nominal p-values < 0.05, these findings cannot be considered statistically reliable without multiple testing correction. Given that ~450,000 CpG sites are tested simultaneously, applying a threshold of p < 0.05 would be expected to yield approximately 22,500 false positives by chance alone, necessitating correction methods

**Suggested Interpretation:** The probe-wise t-test produced results for 485,512 CpG probes. Using nominal p-values, 11,928 probes have `P_Value <= 0.05`; however, these results should be considered preliminary because multiple testing correction has not yet been applied. Since hundreds of thousands of CpG sites are tested simultaneously, nominally significant probes may include false positives, so Bonferroni and Benjamini-Hochberg correction will be needed in Step 10.

## Save Results

```{r save-results}
write.csv(
  t_test_results,
  "results/tables/step09_t_test_results.csv",
  row.names = FALSE
)
```
# Step 10. Multiple Testing Correction



# Step 11. Volcano and Manhattan Plots



# Step 12. Heatmap of the Top 100 CpGs



# Conclusions


