---
title: "Comprehensive Analysis of DNA Methylation Using the Illumina HumanMethylation450 BeadChip Platform"
subtitle: "DNA & RNA Dynamics final project, Team 1. Repository: [Parganini/DRD-Final-Project-Team1](https://github.com/Parganini/DRD-Final-Project-Team1)"
author: "Fatemeh Bayat, Tamime Eter, Seyed Parsa Golshani, Amirhossein Heidari, Mahmoud Ahmed Mohamed Mohamed Ibrahim, Hector Eduardo Parga Najera, Bita Rahmani, Shayan Zanjaniha"
date: "2026-06"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE,
                      fig.align="center", dpi=300, fig.retina=2)
knitr::opts_knit$set(root.dir=normalizePath("."))
```

# Introduction


Team 1 assignments:

- Address for the fluorescence check: `42796479`
- Detection p-value threshold: `0.05`
- Normalization method: `preprocessNoob`
- Differential methylation test: `t-test`

# Step 00. Basic Preparation

Clean the R environment

```{r clean-environment}
rm(list=ls())
```

Load the packages required for this analysis

```{r load-packages}
library(minfi)
library(qqman)
library(gplots)
```

Define the folders used by the project

```{r define-folders}
baseDir <- "data/raw"
rdsDir <- "results/rds"
tablesDir <- "results/tables"
figuresDir <- "results/figures"
```

# Step 01. Data Import and RGChannelSet Creation



# Step 02. Red and Green Fluorescence Channels



# Step 03. Team 1 Address Check



# Step 04. MethylSet Creation



# Step 05. Quality Control



# Step 06. Raw Beta and M Values



# Step 07. Normalization with preprocessNoob



# Step 08. PCA on Normalized Beta Values



# Step 09. Differential Methylation Analysis with t-test



# Step 10. Multiple Testing Correction



# Step 11. Volcano and Manhattan Plots



# Step 12. Heatmap of the Top 100 CpGs



# Conclusions


